Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-22 thesis/_book/_main.tex

@@

            pdfborder={0 0 0},

@@

\title{Conceptual and methodological problems with bias detection and avoidance
in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-06-21}

@@

{
\setcounter{tocdepth}{5}
\tableofcontents
}
\chapter{Introduction}\label{introduction}

Natural language processing (NLP) is a subfield of computer science that
processes and analyzes language in text and speech with the use of
modern programming methods. It has practical applications in everyday
life as it concerns tasks such as email filters, smart assistants,
search results, language translations, text analytics and so on. Models
used to accomplish these tasks need a lot of data to learn from. This
data originates from human activities and historical recordings such as
texts, messages or speeches. It turns out that in the learning process
these models can learn implicit biases that reflect harmful
stereotypical thinking still present in modern societies. There are many
different types of models in NLP depending on a task that they are
supposed to solve. However, all of them need as an input words
represented by means of numbers and this is accomplished with word
embeddings. The word embeddings are usually assigned the values based on
the context in which the words appear. This means that the input data
can have an enormous influence on the outcome. The biases seem to have
their primary source in the way the words are assigned numerical values.
One can find methods that aim at identifying and measuring hidden biases
and/or try to remove them by for example modifying the word embeddings
directly.

One can find various definitions trying to capture what bias and
fairness actually are. With the choice of the definition, implications
into the real-life applications may change as well. Mehrabi, Morstatter,
Saxena, Lerman, \& Galstyan
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
exist different types of biases such as historical bias, representation
bias or measurement bias (the list is long). This indicates how complex
the issue of bias is. Without the proper understanding and awareness of
the problem, people are prone to unconsciously sustain the bias. Mehrabi
et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
distinguish different types of discrimination, some of them will be
briefly described. By protected attributes we mean those qualities,
traits or characteristics that one should not discriminate against.
Direct discrimination occurs when protected attributes of individuals
explicitly result in non-favorable outcomes toward them. Such
discrimination takes place for example when one uses the information
about someone's religious beliefs to evaluate whether they are competent
for the work they apply for. In contrast, in indirect discrimination
individuals appear to be treated equally, but anyway they end up being
treated unjustly due to the hidden effects of biases towards their
protected attributes. An example of such situation is when in a job
advert there is a requirement of 10 years experience instead of the list
of a specific type of experience and knowledge. This job advert can
discriminate indirectly against young people who can have the required
skills, but not yet so much work experience. Systemic discrimination
takes place when policies, customs or behaviors that result from certain
culture or organizational structure lead to discrimination against some
groups of people. Finally, statistical discrimination consists in
relying on group statistics to judge person belonging to a given group.
The topic of discrimination is entangled with another concept; fairness.
It is essential to grasp some concepts of fairness to take them into
consideration while designing implementation of some machine learning
model. In Mehrabi et al.
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
depending on the context and application different definitions may be
applied.

There is a considerable amount of literature available on the topic of
bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
focus on gender biases that may be observable while investigating the
representation of job occupations and gender in terms of their assigned
numerical values. The authors apply cosine similarity measurement to
investigate the phenomenon where (the vectors corresponding to) words
related to jobs that are stereotypically associated with a given gender
are in fact in the model situated closer to this gender. They also use
analogy tasks to evaluate if bias is present in a word embedding model.
They check analogies by comparing pairs of word vectors. For example
they search for the word complementing the puzzle: man is to doctor as
woman is to \ldots{}? First they subtract the word ``man'' from the word
``woman''\footnote{This is pointwise subtraction of vectors, but for the
  simplicity we will often talk about words while really meaning
  vectors. What is meant should be clear from the context.} and then
they search for the ranked list of other words pairs that have similar
vector difference. They also include in the formula a threshold to
ensure that the resulting pairs could not be randomly picked.

However, as Nissim, Noord, \& Goot
(\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are
some limitations to this approach. In practice, most of analogies
implementations do not return any input words. This means that it does
not make sense to expect the algorithm to return the same profession for
both woman and man. Therefore, this method seems to be limited in terms
of bias detection. Other problems are related for example to the choice
of pairs and words that are used to detect the presence of
discrimination, as it is often subjective and without proper
justification. Additionally, the choice of the parameter used in
Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
formula to ensure that word pairs are not picked by random, is also not
justified and changing it drastically influences the results.

@@

Manzini, Lim, Tsvetkov, \& Black
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a
novel way of using cosine similarity to verify if the assumed
resemblance between protected words and harmful attributes exists. They
investigate an approach that enables them to measure the bias for a
class (such as gender, religion or race) and express the final result
with a single metric. Their approach will be analyzed in details in the
next chapter.

Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer
to Implicit Association Test (Greenwald et al., 1998) that measures the
strength of associations between concepts or stereotypes by measuring
human reaction time for special tasks. Humans naturally exhibit some
biases which do not always cause social concern. One can imagine the
intuitive associations between for example insects and flowers, and the
feelings of pleasantness or unpleasantness. In general, people would
rather associate flowers with feeling pleasant than insects, and this
preference in a sense is a bias or prejudice in some direction. However,
this type of preference does not cause an uproar and is a rather morally
neutral case. This example is rather used to show that the methodology
used in IAT makes sense. Islam et al.
(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to
obtain similar results to the ones that measured reaction time with the
use of WEAT metric. Unfortunately, further studies discover other biases
and prejudices that directly influence the quality of other people's
lives and therefore they should be taken care of.

The most common bias detection methods in the literature focus on
comparing the similarity between words from protected groups and those
that are considered to be stereotypical or harmful in some way. One can
find in this group methods such as euclidean distance or cosine
similarity (which is equivalent to dot product if the vectors are
normalized). There are also other ways to detect the effects of biases.
One can investigate the model performance on certain downstream tasks
and validate if the model uses the gender or race information to
complete the task. The currently used methods employing cosine
similarity, often aggregate the similarity values often aggregated in a
way that may lead to hasty conclusions. The averaging of values and the
lack of attention to uncertainty involved in such aggregation may lead
to the incomplete picture of the situation.

@@

Cosine similarity is considered a proper tool for this operation as its
result has a clear connection to geometry and at least for a low number
of dimensions may be easily interpreted. Using this scale, one can
compare vector similarities in a fairly clear manner. When the vectors
are aligned perpendicularly to each other, their similarity equals 0
(which is the same as the cosine of 90 degrees). This tells us that the
similarity between the vectors is small. As the angle between vectors
decreases, cosine similarity approaches one, which stands for the
greatest similarity. It is also possible to obtain negative cosine
similarity, if the value approaches -1 it means that the words are
contrary to each other.

@@

One should note that cosine distance is not exactly a distance measure,
as it does not meet triangle inequality requirements. The triangle
inequality formula says that for any triangle, the sum of the lengths of
any two sides must be greater than or equal to the length of the
remaining side. As shown in
\href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion
in stats.stackexchange.com} in the case of cosine distance it would have
to fulfill this equation 1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) +
cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to
demonstrate that the triangle inequality is not preserved.

@@

Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
similarity between words as the outcome inner product of their
normalized vectors. They focus on examining what the geometry of word
embedding is in regard to ``he'' and ``she'' words. In other words,
whether the similarity between those concepts and other words reflects
expected gender stereotypes. They test this hypothesis by investigating
whether there is a connection between word embeddings representing
certain professions and words referring to gender. They also evaluate
whether automatically produced analogies between words reflect the
stereotypes as well.

@@

A very vivid way to follow their method of arguing that bias in word
embeddings is real is to plot the values of inner product of chosen
words. The plot below does not originate from the original paper but
from
\href{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}{kaggle.com
R notebook using data from GloVe: Global Vectors for Word
Representation}. However, similar visualization may be found there. Data
used to create our plot is as follows.

@@

Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a
different approach towards finding similarities between classes of
words. The authors claim that texts available online are full of direct
or indirect human stereotypes. As a result, word embeddings are prone to
learn and maybe amplify those biases and propagate them further into AI
models that are used for various applications. Cosine distance is used
in the article as a measure to first argue for the existence of
multi-class bias and then to show how through bias mitigation techniques
the bias may be decreased.

@@

In the final step there is an evaluation of the results. The cosine
distances are calculated again but this time using debiased vocabulary.
After taking the mean of all distances one final value is obtained and
then it is compared with the average value from the beginning. If the
cosine distances are on average greater than before then it leads the
authors to the conclusion that improvement has been achieved. As the
cosine distance increases, it is assumed that the association between
protected and stereotypical words decreases.

@@

MAC stand for mean average cosine similarity, although in reality the
the table contains mean distances. What may attract attention is the
fact that the value of cosine distance in ``Biased'' category is already
quite high even before debiasing. High cosine distance indicates low
cosine similarity between values. One could think that average cosine
similarity equal to approximately 0.141 is not significant enough to
consider it as bias. However the authors aim to mitigate ``biases'' in
vectors with such great distance to make it even larger.
Methodologically the question is, on what basis is this small similarity
still considered as a proof of the presence of bias, and whether these
small changes are meaningful. This is in general the problem of scale
and the lack of universal intervals. In contrast, proper intervals
decide whether given cosine similarity is high enough to consider the
words to be more similar than if we chose them at random. We will use
highest posterior density intervals.

@@

We will use the choice of protected words and stereotypical predicates
used in REF. This is a decent point of departure, not only we want to
compare our method to that of REF, but also because this data format is
fairly general (as contrasted, say, with a set up for binary
stereotypes). Note also that the method we develop here can fairly
easily be run for different stereotypization patterns. Let's start with
explaining the method and its deployment using a dataset obtained for
the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we
prepared. In order to obtain this dataset, we calculated the cosine
distance between each protected word and each word from both the
bias-related attribute groups, which were used in the original study,
and to neutral and human control attributes which we added as control
groups. For instance, for religion, the bias-related predicates (coming
from the original study in REF) include muslim bias attributes, jew bias
attributes, christian bias attributes (see a list in the APPENDIX).

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\CommentTok{#head(religion)}
\CommentTok{#library(plyr)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
\CommentTok{#levels(religion$wordClass)}
\CommentTok{#religion$wordClass <- mapvalues(religion$wordClass, }
\CommentTok{#                          from = levels(religion$wordClass),}
\CommentTok{#                          to = c("christian","human","jewish","muslim","neutral"))}
\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}

\CommentTok{#we plug in our visualisation script}
\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
\CommentTok{#two arguments: dataset and protected word}
\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}

\CommentTok{#we plug in our visualisation script}
\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
\CommentTok{#two arguments: dataset and protected word}
\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

We will now think of cosine distance as the output variable, and will
build a few Bayesian models to compare. First, we just build a baseline
model which estimates cosine distance to the attributes separately for
each protected word. The underlying idea is that different protected
words might in general have different relations to all the attributes
and these relations should be our point of departure.

@@

\section{Dataset-level coefficients}\label{dataset-level-coefficients}

@@

We first encountered the GoogleNews word embedding model in Bolukbasi et
al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}). They use it for
their calculations as written in
\href{https://github.com/tolga-b/debiaswe}{Tolga Code}. We decided to
compare the results obtained with the use of Reddit L2 model and the
ones that we got by GoogleNews model. The details of the model can be
found here \href{https://code.google.com/archive/p/word2vec/}{Google
News model}. One of the main differences between the models is that
Reddit word embeggings have 50 dimensions and GoogleNews word embeddings
have 300 dimensions. As the dimension is increasing, the vector can
capture much more information but this is not always the case. According
to some researchers the amount of dimensions should be chosen by taking
into account the type of corpus and its features. In our case both
Reddit and GoogleNews models are already trained and ready to use so we
do not analyse further the choice of their hyperparameters.

@@

It is worth paying attention to the gender debiased dataset. The change
is really minor, still zero out of HPDI range. The \texttt{different}
attributes still have high similarity value. It seems that the distances
for gender protected words are still clustered together which suggests
that this method is not able to catch the complex bias presence. One
could conclude that co-occurrence of certain terms does not always mean
direct connection in terms of terms associations. The fact that some of
stereotypically associated with female attributes still have high
similarity with male protected words suggest that other techniques
should be applied to detect and remove the bias.

@@

In Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is
mentioned that debiasing with this method that is based on bias
component removal may be insufficient. The reason is that words
containing similar biases may still cluster together even after
debiasing.

@@

Bibliography:

>>>>

\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref

@@

            colorlinks=true,
            linkcolor=Maroon,
            citecolor=Blue,
            urlcolor=blue,

@@

\title{Conceptual and methodological problems with bias detection and avoidance
in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-06-22}

@@

{
\hypersetup{linkcolor=black}
\setcounter{tocdepth}{5}
\tableofcontents
}
\chapter{Introduction}\label{introduction}

Natural language processing (NLP) is a subfield of computer science that
processes and analyzes language in text and speech with the use of
modern programming methods. It has practical applications in everyday
life as it concerns tasks such as email filters, smart assistants,
search results, language translations, text analytics and so on. Models
used to accomplish these tasks need a lot of data to learn from. This
data originates from human activities and historical recordings such as
texts, messages or speeches. It turns out that in the learning process
these models can learn implicit biases that reflect harmful
stereotypical thinking still present in modern societies. There are many
different types of models in NLP depending on a task that they are
supposed to solve. However, all of them need as an input words
represented by means of numbers and this is accomplished with word
embeddings. Word embeddings are usually assigned the values based on the
context in which the words appear. This means that the input data can
have an enormous influence on the outcome. One can find methods that aim
at identifying and measuring hidden biases and/or try to remove them by,
for example, modifying the word embeddings directly.

One can find various definitions trying to capture what bias and
fairness actually are. With the choice of the definition, implications
for the real-life applications may change as well. Mehrabi, Morstatter,
Saxena, Lerman, \& Galstyan
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
exist different types of biases such as historical bias, representation
bias or measurement bias (the list is long). This indicates how complex
the issue of bias is. Without the proper understanding and awareness of
the problem, people are prone to unconsciously sustain the bias. Mehrabi
et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
distinguish different types of discrimination, some of them will be
briefly described. By protected attributes we mean those qualities,
traits or characteristics that one should not discriminate against.
Direct discrimination occurs when protected attributes of individuals
explicitly result in non-favorable outcomes toward them. Such
discrimination takes place for example when one uses the information
about someone's religious beliefs to evaluate whether they are competent
for the work they apply for. In contrast, in indirect discrimination
individuals appear to be treated equally, but anyway they end up being
treated unjustly due to the hidden effects of biases towards their
protected attributes. An example of such situation is when in a job
advert there is a requirement of 10 years experience instead of the list
of a specific type of experience and knowledge. This job advert can
discriminate indirectly against young people who can have the required
skills, but not yet so much work experience. Systemic discrimination
takes place when policies, customs or behaviors that result from certain
culture or organizational structure lead to discrimination against some
groups of people. Finally, statistical discrimination consists in
relying on group statistics to judge a person belonging to a given
group. The topic of discrimination is entangled with another concept;
fairness. It is essential to grasp some concepts of fairness to take
them into consideration while designing implementation of some machine
learning model. In Mehrabi et al.
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
depending on the context and application different definitions may be
applied.

There is a considerable amount of literature available on the topic of
bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
focus on gender biases that may be observable in the representation of
job occupations and gender in terms of numerical values. The authors
apply cosine similarity measurement to investigate the phenomenon where
(the vectors corresponding to) words related to jobs that are
stereotypically associated with a given gender are in fact in the model
situated closer to this gender. They also use analogy tasks to evaluate
if a bias is present in a word embedding model. They check analogies by
comparing pairs of word vectors. For example they search for the word
complementing the puzzle: man is to doctor as woman is to \ldots{}?
First they subtract the word ``man'' from the word ``woman''\footnote{This
  is pointwise subtraction of vectors, but for the simplicity we will
  often talk about words while really meaning vectors. What is meant
  should be clear from the context.} and then they search for the ranked
list of other words pairs that have similar vector difference. They also
include in the formula a threshold to ensure that the resulting pairs
could not be randomly picked.

However, as Nissim, Noord, \& Goot
(\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are
some limitations to this approach. In practice, most of analogies
implementations do not return any input words. This means that it does
not make sense to expect the algorithm to return the same profession for
both woman and man. Therefore, this method of bias detection seems
limited. Other problems are related for example to the choice of pairs
and words that are used to detect the presence of discrimination, as it
is often subjective and without proper justification. Additionally, the
choice of the parameter used in Bolukbasi et al.
(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that
word pairs are not picked by random, is also not justified and changing
it drastically influences the results.

@@

Manzini, Lim, Tsvetkov, \& Black
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a
novel way of using cosine similarity to verify if the assumed
resemblance between protected words and harmful attributes exists. They
develop an approach that enables them to measure the bias for a class
(such as gender, religion or race) and express the final result with a
single metric. Their approach will be analyzed in details in the next
chapter.

Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer
to Implicit Association Test (Greenwald et al., 1998) that measures the
strength of associations between concepts or stereotypes by measuring
human reaction time for special tasks. Humans naturally exhibit some
biases which do not always cause social concern. One can imagine the
intuitive associations between for example insects and flowers, and the
feelings of pleasantness or unpleasantness. In general, people would
rather associate flowers than insects with feeling pleasant, and this
preference in a sense is a bias or prejudice in some direction. However,
this type of preference does not cause an uproar and is a rather morally
neutral case. This example is rather used to show that the methodology
used in IAT makes sense. Islam et al.
(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to
obtain similar results to the ones that measured reaction time with the
use of WEAT metric. Unfortunately, further studies discover other biases
and prejudices that directly influence the quality of other people's
lives and therefore they should be taken care of.

The most common bias detection methods in the literature focus on
comparing the similarity between words from protected groups and those
that are considered to be stereotypical or harmful in some way. One can
find in this group methods such as Euclidean distance or cosine
similarity (which is equivalent to dot product if the vectors are
normalized). There are also other ways to detect the effects of biases.
One can investigate the model performance on certain downstream tasks
and validate if the model uses the gender or race information to
complete the task. Unfortunately, the currently used methods employing
cosine similarity, often aggregate the similarity values in a way that
may lead to hasty conclusions. The averaging of values and the lack of
attention to uncertainty involved in such aggregation may lead to an
incomplete picture of the situation.

@@

Cosine similarity is considered a proper tool for this operation as its
result has a clear connection to geometry and at least for a low number
of dimensions may be easily interpreted. Using this scale, one can
compare vector similarities in a fairly clear manner. When the vectors
are aligned perpendicularly to each other, their similarity equals 0
(which is the same as the cosine of 90 degrees). This tells us that the
similarity between the vectors is small. As the angle between vectors
decreases, cosine similarity approaches one, which stands for the
greatest similarity. It is also possible to obtain negative cosine
similarity. If the value approaches -1, this intuitively means that the
words are contrary to each other.

@@

One should note that cosine distance is not exactly a distance measure,
as it does not meet triangle inequality requirements. The triangle
inequality formula says that for any triangle, the sum of the lengths of
any two sides must be greater than or equal to the length of the
remaining side. As shown in
\href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion
in stats.stackexchange.com}\footnote{\url{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}}
in the case of cosine distance it would have to fulfill this equation
1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses
specific unit vectors it is easy to demonstrate that the triangle
inequality is not preserved.

@@

Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
similarity between words as the outcome inner product of their
normalized vectors. They focus on examining what the geometry of a word
embedding is in regard to ``he'' and ``she'' words. In other words,
whether the similarity between those concepts and other words reflects
expected gender stereotypes. They test this hypothesis by investigating
whether there is a connection between word embeddings representing
certain professions and words referring to gender. They also evaluate
whether automatically produced analogies between words reflect the
stereotypes as well.

@@

A very vivid way to follow their method of arguing that bias in word
embeddings is real is to plot the values of inner product of chosen
words. The plot below does not originate from the original paper but
from
\href{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}{kaggle.com
R notebook using data from GloVe: Global Vectors for Word
Representation}.\footnote{\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}.}
However, similar visualization may be found there. Data used to create
our plot is as follows.

@@

Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a
different approach towards finding similarities between classes of
words. Cosine distance is used in the article as a measure to first
argue for the existence of multi-class bias and then to show how through
bias mitigation techniques the bias may be decreased.

@@

In the final step they evaluate the results. The cosine distances are
calculated again but this time using the debiased vocabulary. After
taking the mean of all distances one final value is obtained and then it
is compared with the average value from the beginning. If the cosine
distances are on average greater than before then it leads the authors
to the conclusion that improvement has been achieved. As the cosine
distance increases, it is assumed that the association between protected
and stereotypical words decreases.

\newpage

@@

MAC stand for mean average cosine similarity, although in reality the
the table contains mean distances. What may attract attention is the
fact that the value of cosine distance in ``Biased'' category is already
quite high even before debiasing. High cosine distance indicates low
cosine similarity between values. One could think that average cosine
similarity equal to approximately 0.141 is not significant enough to
consider it as bias. However the authors aim to mitigate ``biases'' in
vectors with such great distance to make it even larger.
Methodologically the question is, on what basis is this small similarity
still considered as a proof of the presence of bias, and whether these
small changes are meaningful. This is in general the problem of scale
and the lack of universal intervals. In contrast, statistical intervals
will help us decide whether a given cosine similarity is high enough to
consider the words to be more similar than if we chose them at random.
We will use highest posterior density intervals, in line with Bayesian
methodology.

@@

We will use the choice of protected words and stereotypical predicates
used in Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). This is a
decent point of departure, as not only we want to compare our method to
that of Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}), but also
because this data format is fairly general (as contrasted, say, with a
set up for binary stereotypes). Note also that the method we develop
here can fairly easily be run for different stereotypization patterns.
Let's start with explaining the method and its deployment using a
dataset obtained for the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we
prepared. In order to obtain this dataset, we calculated the cosine
distance between each protected word and each word from both the
bias-related attribute groups, which were used in the original study,
and to neutral and human control attributes which we added as control
groups. For instance, for religion, the bias-related predicates (coming
from the original study in Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})) include
muslim bias attributes, jew bias attributes, christian bias attributes
(see a list in the Appendix).

@@

\pagebreak  \vspace{1mm} \footnotesize

@@

@@

@@

@@

We will now think of cosine distance as the output variable, and will
build a few Bayesian models to compare. First, we just build a baseline
model which estimates cosine distance to the attributes separately for
each protected word. The underlying idea is that different protected
words might in general have different relations to all the attributes
and these relations should be our point of departure.\footnote{The
  construction of the Bayesian models and code for visualisations is due
  to Rafal Urbaniak.}

@@

\section{\texorpdfstring{Dataset-level coefficients
\label{sec:datasetsLevelCoeffs}}{Dataset-level coefficients }}\label{dataset-level-coefficients}

@@

We first encountered the GoogleNews word embedding model in Bolukbasi et
al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}). They use it for
their calculations as written in
\href{https://github.com/tolga-b/debiaswe}{Tolga Bolukbasi
code}\footnote{\url{https://github.com/tolga-b/debiaswe}}. We decided to
compare the results obtained with the use of Reddit L2 model and the
ones that we got by GoogleNews model. The details of the model can be
found here \href{https://code.google.com/archive/p/word2vec/}{Google
News model}. One of the main differences between the models is that
Reddit word embeggings have 50 dimensions and GoogleNews word embeddings
have 300 dimensions. As the dimension increases, the vectors can capture
much more information, but this is not always the case. According to
some researchers, the amount of dimensions should be chosen by taking
into account the type of corpus and its features. In our case both
Reddit and GoogleNews models are already trained and ready to use so we
do not analyse further the choice of their hyperparameters.

@@

It is worth paying attention to the gender debiased dataset. The change
is really minor, still with zero out of the HPDI range. The
\texttt{different} attributes still have high similarity value. It seems
that the distances for gender protected words are still clustered
together which suggests that this method is unable to catch the complex
bias presence. One could conclude that co-occurrence of certain terms
does not always mean direct connection in terms of terms associations.
The fact that some of stereotypically associated with female attributes
still have high similarity with male protected words suggest that other
techniques should be applied to detect and remove the bias.

@@

@@

\chapter*{Appendix}\label{appendix}
\addcontentsline{toc}{chapter}{Appendix}
